Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-17 paper.tex
First attempt at rewriting barrier section
Also clarify that reduce-then-scan relies on pipeline barriers.

I think more work is needed to really clearly explain the memory model requirements and reliance on forward progress guarantees, but hopefully this is a step in the right direction.

\subsubsection{Inter-Workgroup Barriers}
The virtualization of processors introduces an emergent high-level limitation: the absence of an inter-workgroup barrier. As previously mentioned, workgroups cannot context switch. Although contemporary GPU hardware~\cite{} supports workgroup preemption for prioritizing real-time tasks or managing multi-process workloads, this functionality is deliberately excluded from the GPU programming model. This stems from the prohibitively high overhead in latency and storage associated with moving workgroup contexts on and off chip. In the oversubscription paradigm, where the number of workgroups is theoretically unbounded and typically far exceeds the available multiprocessors, every invocation of such a barrier would necessitate context-switching between all launched workgroups. Due to this high cost, no GPU programming environment furnishes developers with a true unbounded-launch inter-workgroup barrier.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} Instead, when inter-workgroup synchronization is necessary, developers must rely on kernel launches as synchronization points. However, kernel launches are suboptimal as barriers: they incur overheads due to the heterogeneous nature of the operation, and all execution context is lost between launches.

\raph{I think the above is conflating memory barriers with control barriers. All modern GPU shading languages (since Metal 3.2) provide memory barriers with inter-workgroup (device) scope, but the above paragraph makes more sense interpreted as a control barrier. It might also be worth mentioning that kernel launches do not necessarily sychronize; in Vulkan it's really the pipeline barrier that does the synchronization. I can take a whack at re-writing this.}

@@

\subsection{Ideal Scan Architecture}
\emph{Chained Scan} and \emph{Reduce-Then-Scan} represent two contrasting paradigms in scan design. By avoiding explicit barrier-based inter-workgroup synchronization, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler. In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, requiring fixed launches and explicit occupancy control, forcing developers to manually tune launch sizes or implement occupancy discovery techniques.

@@

\raph{I'm having trouble with the above paragraph. By my read, Reduce-Then-Scan avoids inter-workgroup synchronization other than the pipeline barriers between kernel launches. Chained-Scan and Decoupled-Lookback both rely on inter-workgroup synchronization. Not sure why the "in some cases" qualifier either; seems to me you need synchronization except in the trivial case the entire array fits in one workgroup.}

\raph{"Fixed launch" means launching a fixed number of workgroups and having each workgroup process multiple blocks, generally total blocks / num workgroups? I think I get the argument here but it's more subtle and not an inherent problem with RTS. You make a choice: either you do fixed launch and have to tune, *or* you can do a fixed number of blocks per workgroup, but in the latter case you need to need vary the number of dispatches based on total number of blocks. That's annoying but not a showstopper.}

\thomas{Edited this section in response to above comments. Distinguished Chained Scan as specifically eliding barrier based sychronization. Linked fixed launch more clearly to manual tuning and occupancy discovery. Shortened section. OK to remove comments?}

@@

  $\textit{local\_vectors}$ \gets $\textnormal{array}\langle \textnormal{vec4}\langle u32\rangle, \textit{p} \rangle$\;

>>>>

\subsubsection{Inter-Workgroup Barriers}
Parallel programs in general require synchronization between multiple threads (those that don't are considered embarrassingly parallel). A standard technique for synchronization is \emph{barriers.} GPUs provide a limited palette of barrier primitives. At the coarsest granularity, \emph{pipeline barriers} fully synchronize both control and memory between multiple kernel launches. However, pipeline barriers are extremely expensive and efficient algorithms strictly minimize their use. At a much finer granularity, \emph{workgroup barriers} provide reasonably efficient synchronization between the threads in a workgroup. However, synchronization between workgroups is more limited. Standard GPU programming environments provide no control barrier between workgroups.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} The availability of a memory-only barrier varies across GPU environments; it is present in Direct3D and Vulkan, but was missing in versions of Metal prior to 3.2. Because of its non-portability, it is also absent from the WebGPU specification. Even when available, the use of such memory barriers causes performance issues as they are usually implemented as a full cache flush.

@@

\subsection{Ideal Scan Architecture}
\emph{Chained Scan} and \emph{Reduce-Then-Scan} represent two contrasting paradigms in scan design. By avoiding explicit pipeline barrier based inter-workgroup synchronization, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler. In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, requiring fixed launches and explicit occupancy control, forcing developers to manually tune launch sizes or implement occupancy discovery techniques.

@@

@@

  $\textit{local\_vectors} \gets \textnormal{array}\langle \textnormal{vec4}\langle u32\rangle, \textit{p} \rangle$\;
